`plot.dads`

### 2023-03-27 inst/gitbook/06_plotting_dads_objects.Rmd
Finished plotting section

---
title: "`plot.dads`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```

@@

```{r}
## Playing with some more options
plot(my_tree, main = "A tree and traits",
     ## Changing nodes colours and type
     col = c(tips = "pink", nodes = "purple"), pch = 21,
     ## Changing edges colour and type
     lwd = 2, lty = 3, edges = "yellow",
     ## Changing the x axis orientation and label
     xlim = c(0, 2.5), xlab = "Time into the past")
```

@@

```{r}
## Adding the default legend
plot(my_tree, legend = TRUE)
```

```{r, eval = FALSE}
plot(my_data, trait = 2, col = c("purple", "blue"))

## Specifying a 3D trait process
my_3D_trait <- make.traits(n = 3)
## Simulating a birth death tree with that 3D trait
my_data <- dads(bd.params = list(extinction = 0.2),
                stop.rule = list(max.living = 50),
                traits    = my_3D_trait)

## Plotting the 
plot(my_data, trait = c(1, 3), col = terrain.colors, tips.nodes = "black", legend = TRUE)
```

<!-- 
test_that("plot.dads works for traits", {

    list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
    ## Right output
    expect_null(plot.dads(list_of_traits))
    expect_null(plot.dads(list_of_traits, trait = 2))
    expect_null(plot.dads(list_of_traits, trait = 3))
})

test_that("plot.dads works for dads", {

    ## Simple example    
    test <- dads(stop.rule = list("max.living" = 10), traits = make.traits())
    ## Default plot
    expect_null(plot(test))
    ## Other options
    expect_null(plot(test, main = "what?", xlim = c(0, test$tree$root.time), col = c(tips = "pink", nodes = "red", edges = "purple"), ylab = "data"))

    ## A more complex trait
    complex_traits <- make.traits(process = c(BM.process, BM.process), n = c(2,3), process.args = list(list(Sigma = diag(2)), list(Sigma = matrix(1/3, 3, 3))), trait.names = c("bib", "bob"))
    test <- dads(stop.rule = list("max.living" = 10), traits = complex_traits)
    expect_null(plot(test, trait = 3))

    ## A big tree
    set.seed(12)
    big_test <- dads(bd.params = list(speciation = 1, extinction = 1/3), stop.rule = list(max.taxa = 1000), traits = make.traits())
    expect_null(plot(big_test, cex = 0.5))

    list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
    ## Right output
    expect_null(plot.dads(list_of_traits))
    expect_null(plot.dads(list_of_traits, trait = 2))
    expect_null(plot.dads(list_of_traits, trait = 3))
})

test_that("example works", {
    ## Specifying a trait process
    my_trait <- make.traits()
    ## Plotting a trait process
    expect_null(plot(my_trait, main = "A Brownian Motion"))

    ## Simulating a tree with ten taxa
    my_tree <- dads(stop.rule = list(max.taxa = 10))
    ## Plotting a simple birth death tree (using ape::plot.phylo)
    expect_is(plot(my_tree, main = "A pure birth tree"), "list")

    ## Simulating a tree with traits
    my_data <- dads(stop.rule = list(max.taxa = 10),
                    traits    = my_trait)
    ## Plotting the tree and traits
    expect_null(plot(my_data))

    ## Specifying a 3D trait process
    my_3D_trait <- make.traits(n = 3)
    ## Simulating a birth death tree with that trait
    my_data <- dads(bd.params = list(extinction = 0.2),
                    stop.rule = list(max.living = 50),
                    traits    = my_3D_trait)

    ## Plotting the second trait and the tree (default)
    ## The colours are purple for nodes and blue for tips
    expect_null(plot(my_data, trait = 2, col = c("purple", "blue"),
         edges = "pink", tips.nodes = "black"))

    ## Plotting the first and third trait correlation
    ## The colours are a heat map based on the elements age
    expect_null(plot(my_data, trait = c(1, 3), col = terrain.colors,
         edges = "grey", tips.nodes = "black"))

    ## Plotting the first and third trait correlation in 3D
    expect_null(plot(my_data, trait = c(1,3), col = rainbow,
         edges = "grey", tips.nodes = "black", use.3D = TRUE))

    ## Plotting all traits in 3D (without branch lengths)
    expect_null(plot(my_data, trait = c(1:3), col = heat.colors,
         edges = "grey", tips.nodes = "black", use.3D = TRUE, type = "s", radius = 0.1))
}) -->

>>>>

---
title: "`plot.dads`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
# bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(dads)
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
options(rgl.useNULL = TRUE) # Suppress the separate window.
```

@@

```{r, fig.height = 6, fig.width = 6}
## Playing with some more options
plot(my_tree, main = "A tree and traits",
     ## Changing nodes colours and type
     col = c(tips = "pink", nodes = "purple"), pch = 21,
     ## Changing edges colour and type
     lwd = 2, lty = 3, edges = "yellow",
     ## Changing the x axis orientation and label
     xlim = c(0, 2.5), xlab = "Time into the past")
```

@@

```{r, fig.height = 6, fig.width = 6}
## Adding the default legend
plot(my_tree, legend = TRUE)
```

## 3D version!

Of course, if you're simulating multiple traits, you can always plot different ones.

```{r, fig.height = 6, fig.width = 6}
## Specifying a 3D trait process
my_3D_trait <- make.traits(n = 3, )
## Simulating a birth death tree with that 3D trait
my_data <- dads(bd.params = list(extinction = 0.2),
                stop.rule = list(max.living = 50),
                traits    = my_3D_trait)
```

You can toggle which trait to plot using the `trait` option.
Either by providing a single value to plot that specific trait against time or by providing two traits.

```{r, fig.height = 12, fig.width = 6}
par(mfrow = c(2,1))
## Plotting the second trait
plot(my_data, trait = 2, main = "Trait 2")
## Plotting the correlation between trait 1 and 2
plot(my_data, trait = c(1,2), main = "Traits 1 and 2")
```

As mentioned above, the `col` argument can take a function for scaling the elements colours with time.
This can be useful for adding time a third dimension to these 2D plots:

```{r}
## Plotting the correlation between trait 1 and 2
## with time as a 3rd dimensions
plot(my_data, trait = c(1,2), main = "Traits 1 and 2",
     col = grDevices::heat.colors, legend = TRUE,
     ## Highlighting the tips in black for visibility
     tips.nodes = "black")
```

But there's more! You can also plot these results using actual 3D plots that you can spin and all that!
This is done through the `rgl` package and activated in `dads` using the option `use.3D = TRUE`.

If two traits are provided, the 3rd dimensions is going to be time by default:

```{r}
## Plotting the tree and 2 traits in 3D: woah!
plot(my_data, trait = c(1, 2), use.3D = TRUE)
rglwidget()
```

If three traits are provide, you can set up a fourth coloured dimension as time (similarly to the 2D plots with the colour gradients).
This uses the `rgl` functions, namely `lines3d`, `points3d` (if `type = "p"`; default) or `sphere3d` (if `type = "s"`).
The additional arguments `...` are directly handled and attributed to the corresponding function.

```{r}
## Plotting the tree and 3 traits in 3D
plot(my_data, trait = c(1, 2, 3), use.3D = TRUE,
     col = grDevices::heat.colors, type = "s", radius = 0.1)
rglwidget()
```
